---
title: "machine_failure"
author: "Srihari Seshadri"
date: "June 2, 2019"
output: html_document
---

```{r warning=FALSE, message=FALSE}
library(dplyr)
# Import the relevant libraries
suppressWarnings(library(plyr))
suppressWarnings(library(tseries))
suppressWarnings(library(forecast))
suppressWarnings(library(lmtest))
suppressWarnings(library(car))
suppressWarnings(library(arfima))
suppressWarnings(library(tsibble))
suppressWarnings(library(ggplot2))
suppressWarnings(library(zoo))
```

```{r}
mf.tsdata <- read.csv("./machine_temperature_system_failure.csv")
head(mf.tsdata)
```

```{r}
plot.ts(mf.tsdata$value)
```


```{r}
acf(mf.tsdata$value, main="ACF plot")
```

# Test for stationarity

### KPSS test
(KPSS) tests are used for testing a null hypothesis that an observable time series is stationary
```{r}
kpss.test(mf.tsdata$value)
```

This indicates that the TS is NOT stationary

### ADF Test
Test for the null hypothesis of a unit root of a univarate time series x (equivalently, x is a non-stationary time series).
```{r}
adf.test(mf.tsdata$value, alternative="stationary", k=10)
```
Contradicts KPSS Test. Indicates that the time series IS Stationary

(https://stats.stackexchange.com/a/235916/102752)
Unit root test: you can't reject H0; KPSS test: reject H0. Both imply that series has unit root.


# Test for Seasonality

```{r}
mf_ma = ts(na.omit(mf.tsdata$value), 
              frequency=300)
decomp = stl(mf_ma, s.window="periodic")
mf.tsdata$value_deseas <- seasadj(decomp)
plot(decomp)

# Recheck for stationarity
kpss.test(mf.tsdata$value_deseas)

adf.test(mf.tsdata$value_deseas, alternative="stationary", k=10)

```

The results haven't changed. ADF still indicates stationarity

# Test for Heteroskedasticity and STationarity

We can use the Phillips–Perron test to test for Stationarity while accounting for heteroskedasticity.

https://en.wikipedia.org/wiki/Phillips%E2%80%93Perron_test
https://rdrr.io/cran/aTSA/man/pp.test.html

The test is similar to ADF in terms of hypo testing. i.e., it's a unit root test, alternate hypo is stationarity

```{r}
pp.test(mf.tsdata$value)
```


The PP test also indicates stationarity. 



