---
title: "NYC EDA"
output: html_notebook
---


```{r warning=FALSE, message=FALSE, echo=FALSE}
library(dplyr)
# Import the relevant libraries
#suppressWarnings(library(plyr))
suppressWarnings(library(tseries))
suppressWarnings(library(forecast))
suppressWarnings(library(lmtest))
#suppressWarnings(library(car))
#suppressWarnings(library(arfima))
#suppressWarnings(library(tsibble))
suppressWarnings(library(ggplot2))
suppressWarnings(library(gridExtra))
#suppressWarnings(library(zoo))
```

### Have a look at the data

We have 7 months of data, from 2014/07/01 till 2015/01/31.

NOTE : Scale data to bring to manageable levels. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
nyctaxi.data <- read.csv("./nyc_taxi.csv")
head(nyctaxi.data)
tail(nyctaxi.data)
dim(nyctaxi.data)
nyctaxi.data$timestamp_dt <- as.POSIXct(nyctaxi.data$timestamp)
#nyctaxi.data$value_boxcox <- BoxCox(nyctaxi.data$value, lambda = 3.5)
#head(nyctaxi.data)

plot1 <- ggplot(nyctaxi.data, aes(x = value)) + geom_histogram() + theme_light()
plot2 <- ggplot(nyctaxi.data, aes(sample = value)) +  geom_line(mapping=aes(x=timestamp_dt, y=value)) + theme_light()
grid.arrange(plot1, plot2, nrow=3)

acf(nyctaxi.data$value)

```

Plot the graphs for just one week

```{r message=FALSE, warning=FALSE, echo=FALSE}
subsamp <- nyctaxi.data[nyctaxi.data$timestamp_dt < as.POSIXct("2014-07-08 00:00:00"),]
plot1 <- ggplot(subsamp, aes(x = value)) + geom_histogram() + theme_light()
plot2 <- ggplot(subsamp, aes(sample = value)) +  geom_line(mapping=aes(x=timestamp_dt, y=value)) + theme_light()
grid.arrange(plot1, plot2, nrow=2)
```

----------
### Visualise the anomalies


```{r warning=FALSE, message=FALSE, echo=FALSE}
anomaly_points <- as.POSIXct(c(
  "2014-11-01 19:00:00",
  "2014-11-27 15:30:00",
  "2014-12-25 15:00:00",
  "2015-01-01 01:00:00",
  "2015-01-27 00:00:00"
))

nyctaxi.data$flag_anomaly <- ifelse(nyctaxi.data$timestamp_dt %in% anomaly_points, TRUE, FALSE)

nyctaxi.data %>% filter(timestamp_dt %in% anomaly_points)

#ggsave(filename="taxi_anomalies.pdf", w=20, h=5, dpi=500)

ggplot() + 
  geom_line(mapping=aes(x=timestamp_dt, y=value), data=nyctaxi.data) + 
  geom_vline(xintercept=as.numeric(anomaly_points), linetype="dotted", color="red", size=1) +
  theme_light()
```


----------

## Tests for Stationarity

```{r warning=FALSE, message=FALSE, echo=FALSE}
# Recheck for stationarity
kpss.test(nyctaxi.data$value)
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
adf.test(nyctaxi.data$value, alternative="stationary", k=10)
```

The signal is definitely stationary. 

##### Test for Heteroskedasticity and Stationarity

We can use the Phillips–Perron test to test for Stationarity while accounting for heteroskedasticity.

https://en.wikipedia.org/wiki/Phillips%E2%80%93Perron_test
https://rdrr.io/cran/aTSA/man/pp.test.html

The test is similar to ADF in terms of hypo testing. i.e., it's a unit root test, alternate hypo is stationarity

```{r warning=FALSE, message=FALSE, echo=FALSE}
pp.test(nyctaxi.data$value)
```

------------------------------------------------------------------------------------
------------------------------------------------------------------------------------
------------------------------------------------------------------------------------

# Test and compensate for Seasonality and Trend

Frequency is the expected repitition window. 

Trial and error, getting order of Moving Average as 2 months of time, i.e., 24 * 2 * 7

```{r echo=FALSE, message=FALSE, warning=FALSE}
ma.order <- 24*2*7
```


```{r warning=FALSE, message=FALSE, echo=FALSE}
trend_nyc = ma(nyctaxi.data$value, order = ma.order, centre = T)

ggplot() +
  geom_line(mapping=aes(x=timestamp_dt, y=value), color="black", data=nyctaxi.data) + 
  geom_line(mapping=aes(x=nyctaxi.data$timestamp_dt, y=trend_nyc), color="red") 
```

```{r warning=FALSE, message=FALSE, echo=FALSE}

nyc_ma = ts(na.omit(nyctaxi.data$value), 
              frequency=ma.order)
decomp = stl(nyc_ma, s.window="periodic")
nyctaxi.data$value_deseas <- seasadj(decomp)
plot(decomp)
```

Eyeballing the graph -> it seems to be an additive model, i.e., data = seasonal + trend + remainder

```{r warning=FALSE, message=FALSE, echo=FALSE}
acf(nyctaxi.data$value_deseas)
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
ggsave(filename="taxi_anomalies.pdf", w=20, h=5, dpi=500)

ggplot() + 
  geom_line(mapping=aes(x=timestamp_dt, y=value_deseas), color="black", data=nyctaxi.data) + 
  geom_line(mapping=aes(x=timestamp_dt, y=value), color="grey", alpha=0.5, data=nyctaxi.data) + 
  geom_vline(xintercept=as.numeric(anomaly_points), linetype="dotted", color="red", size=1) +
  theme_light()
  
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
kpss.test(nyctaxi.data$value_deseas)

adf.test(nyctaxi.data$value_deseas, alternative="stationary", k=10)

pp.test(nyctaxi.data$value)
```


-------------

# Modelling the data

